---
title: "Ontology and Corpus Aided Searching of Complex Data Resources"
author: "RoGe, VC, RaGo"
date: "2023-06-23"
output: 
  bookdown::html_document2: 
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Ontology and Corpus Aided Searching of Complex Data Resources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

bibliography: references.bib

link-citations: true

format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("gwasCatSearch")
library("kableExtra")
library("DT")
```

## Introduction


Larger and more complex biomedical data sets are becoming increasingly
available and they provide substantial opportunities for discovery.
Realizing the potential for discovery is challenged by the complexity of
the data, the inherent relationships between phenotypes and the need for
data analysts to be able to navigate some of the complexities of both
biological and medical terminologies. In this paper we explore using
corpus search tools (as exemplified in the `corpustools` package
@atteveldtCorpustoolsManagingQuerying2023) to help search through the set of traits that are
reported by the GWAS catalog. These traits have been hand curated by the
EBI and mapped to terms in the Experimental Factor Ontology (EFO,
@maloneModelingSampleVariables2010). However, the tools available for searching these data are
mainly variants of regular expression matching, which as very limited
expressivity for the complex nature of medical and biological terms.

The GWAS Catalog (@sollisNHGRIEBIGWASCatalog2023) reports on a set of curated GWAS
association studies. Since 2010, delivery and development of the Catalog
has been a collaborative project between the EMBL-EBI and NHGRI. As of
2024-01-30, the GWAS Catalog contains 6715 publications, 571148 top
associations and 67322 full summary statistics. GWAS Catalog data is
currently mapped to Genome Assembly GRCh38.p14 and dbSNP Build 156.

Among the curation activities is a process of mapping the traits that
are studied to the EFO (@maloneModelingSampleVariables2010). In part this allows one to search for
studies about a particular disease, but for a variety of reasons this
search can be quite complicated. Here we report on tools and
approaches that can be used to help find GWAS studies according to
specific criteria. We create two searchable corpora, one that is EFO
centric and the other that is GWAS trait centric. There are a number of 
anomalies in both the EFO and the GWAS catalog that make it essential that
we have effective tools for cross-referencing searches. Our software is contained in the `gwasCatSearch` package (@gentlemanGwasCatSearchPackageSupport2024a) 
which is open source and freely available.

## Methods

### Data
We obtain data from the GWAS Catalog and transform it into a set of tables using tools in the GWASCatalogSearchDB (https://github.com/ccb-hms/GWASCatalogSearchDB). This processing produces
a number of tables that can be accessed via a SQL database; we use SQLite (@sqlite2020hipp). These tables include the raw data from the EBI, several tables organized around the Experimental Factor Ontology (EFO), the Uber-anatomy ontology (UBERON @Uberon2012 ).
An ontology is a set of concepts and categories that describe a subject area or domain of knowledge that expresses their properties and the relations between them. Ontologies form one of the bases for knowledge engineering and representation and are important in understanding and referencing biomedical entities. One of the challenges that needs to be addressed is the fact that the same concept might usefully appear in different ontologies and users will need tools to help identify similar concepts. A second challenge is that to be useful the ontologies must be accessible and typically that means using the web or a database implementation. There are two useful constructs that help with this, the compact uniform resource identifiers (CURIEs), and the internationalized resource identifiers (IRIs)  which in ontology space are used as *names* terms in the ontology. A CURIE has three parts: a prefix; a delimiter; and a unique local (to the ontology described by the prefix) identifier from the given nomenclature. The CURIE EFO:0005305 which represents the concept *atrioventricular node disease* has a prefix of EFO, a delimiter of a colon (:) and a unique local identifier of 0005305.  The corresponding IRI for that term is http://www.ebi.ac.uk/efo/EFO_0005305, which gives a web location that can be examined by a user.

The disease location information was extracted directly from EFO statements of the form:  X has_disease_location Y, where Y is typically an UBERON term representing an anatomical location. If a term does not have an explicitly stated location relationship, we determine if it has an inferred location. This is done by recursively checking if a parent in the ontology hierarchy has a location. The first location found is used, and if none is found then it is marked as missing. 
For example, *bronchitis*  (EFO:0009661) does not have an explicitly stated location in EFO, but it inherits one from its immediate parent (*bronchial disease*), which has location *bronchus*, (UBERON:0002185).

As part of the processing we construct two dataframes, one based on the EFO terms `efo_df` and the other based on the GWAS catalog studies `gwc_df`. These are described in Table 1. We create two different corpora for searching. One based on the EFO and
the relationships inherent in it, `efo_tc`, and the second based on the GWAS
catalog studies, `gwc_tc`.

The EFO dataframe has one row for each term in the EFO ontology and columns labeled `Subject`: the corresponding EFO CURIE; `Object`: the textual description of the trait; `IRI`: the IRI for that term; `DiseaseLocation`: the UBERON CURIE for the location of the `trait` if there is one; `Direct`: the number of GWAS catalog studies that were mapped directly to that trait; `Inherited`: the number of traits that were mapped to terms that are more specific (according to the ontology) than the `Object`.

Since a GWAS study can be directly mapped to more than one EFO term we have made our
dataframe have one row for each unique combination of GWAS catalog study ID and EFO term.
The GWAS catalog dataframe has one row for each GWAS catalog study and columns labeled `STUDY.ACCESSION`: the GWAS catalog study identifier; `DISEASE.TRAIT`: the GWAS catalog text description of the trait; `MAPPED_TRAIT`: the textual description of the trait from EFO that this study was *directly* mapped to; `MAPPED_TRAIT_URI`: the URI for the mapped trait; `MAPPED_TRAIT_CURIE`: the CURIE for the mapped trait.

Each GWAS study has a specific trait that was studied and that is
reported as a text string. That trait is mapped to a specific term in
the EFO ontology, and we refer to those as **direct** mappings. The terms in the EFO ontology
are related to each other via a directed acyclic graph. By construction there is a single root node. More specific terms, such as the relationship between cancer and pancreatic cancer are encoded with pancreatic cancer being the **child** of cancer,
while cancer is the **parent** of pancreatic cancer.  The descriptions can become more refined, and we define the **descendents** of a node to be the set of child nodes, where one recurses down to the point where a node has no children.  Such a node is referred to as a *leaf*.  Similarly **ancesters** are the parents, again applied recursively to the root node of the ontology.  Our software contains tools to obtain the children, descendents, parents, and ancestors as well as to perform other operations on the ontology. 

Tools for representing and organizing biomedical entities and concepts are key components of
any system that process disease, medical or epidemiological data.  There are many challenges among them is the need to have some way to have two different systems align on some concepts. On the seminatic web Uniform Resource Identifiers (URIs) are used to reference named entities. So URI basically give a web address where more details on the concept can be found.

Given the structure of the ontology we can introduce the concept of an *inherited* mapping. A mapping at a node E is inherited if it was a direct mapping to some descendant of E.  Basically this is
providing a way to find, and hence examine, the EFO terms that map to more specific versions of an EFO term.  So, while we might want to know the GWAS catalog traits that map to cancer, we may also want to find the traits that map to more specific types of cancer.  This concept allows us to do that in a rigorous way.

```{r ontoHelperFuns, echo=FALSE}

##a leaf is a node that has no children
##the root is a node that has no parents

isLeaf = function(x) {
  descend = EFOdescendants(x)
  sp = split(descend, names(descend))
  ans = sapply(sp, function(x) length(x) < 2)
  ans
}

t1 = isLeaf(c('MONDO:0019640','EFO:0004194'))

z2 = EFOdescendants("EFO:0005741")
z3 = EFOancestors("EFO:0005741")

isRoot = function(x) {
  parents=EFOancestors(x)
  sp = split(parents, names(parents))
  ans = sapply( sp, function(x) length(x) < 2)
  ans
}
```
### Some important characteristics of the GWAS Catalog

Here we explore some of the characteristics of the Catalog in order to facilitate some of the discussion in the remainder of the paper.


```{r cmpTableofMappingCounts, echo=FALSE }
pp = table(gwc_df$STUDY.ACCESSION)
pps = sort(pp,dec=T)[1:10]
ptab = table(pp)
ptab = c(ptab[1:9], sum(ptab[10:length(ptab)]))
names(ptab)[10] = "10+"
kable_styling(kbl(t(ptab), caption="Number of EFO Studies by EFO terms per Study"), bootstrap_options = "striped", full_width = FALSE)
```


There are `r length(unique(gwc_df$STUDY.ACCESSION))` studies reported.  Each study can be mapped to one or more EFO terms. We counted the number of EFO terms each study was mapped to and report those in Table \@ref(tab:cmpTableofMappingCounts). The counts range from 1 to `r max(ptab)` with the study `r names(pp)[which.max(pp)]` having the most.

```{r studiesperPMID, echo=FALSE}
xx = DBI::dbGetQuery(gwasCatSearch_dbconn(), "SELECT PUBMEDID, \"STUDY.ACCESSION\" FROM gwascatalog_metadata")
tt = sort(table(xx$PUBMEDID),dec=T)
```

The GWAS catalog also provides a table with one entry per STUDY.ACCESSION number and other variables such as the PUBMEDID for the associated publication. We next examine how many GWAS Catalog studies are associated with different papers. We find that the paper associated with PMID `r names(tt)[1]` has `r tt[1]` studies associated with it.  And there are in fact `r sum(tt > 100)` publications which have more than 100 GWAS studies associated with them.  This seems hard to understand and any analysis that includes these studies will require the authors to dig quite deeply into the methods used by the GWAS Catalog.

```{r grepCKD, echo=FALSE}
xx = DBI::dbGetQuery(gwasCatSearch_dbconn(), "SELECT PUBMEDID, \"DISEASE.TRAIT\" FROM gwascatalog_metadata")
gg = grepl("chronic kidney disease", xx$DISEASE.TRAIT, ignore.case = T)
PMs = xx$PUBMEDID[gg]
```

We last point out that assigned disease traits, the DISEASE.TRAIT column, in the GWAS Catalog metadata may also cause some challenges.  When we searched all disease trait strings for any occurance of the string *chronic kidney disease*, ignoring capitalization, and found `r sum(gg)` matches.  These correspond to `r length(unique(PMs))` PUBMEDIDs.  

### A search corpus based on EFO

To create a searchable EFO-based corpus we think of
each EFO term as a *document*. The EFO provides a text description of
the term as well as a set of known synonyms. We show the EFO provided synonyms for EFO:0003086 in Table \@ref(tab:exampleEFO). We expand the document by including the GWAS trait label
for any GWAS study that has been mapped **directly** to that EFO term.  Our rationale
is that the curators assigned that trait description to the specific EFO term, so it 
is essentially, in their eyes, a synonym.


```{r exampleEFO, echo=FALSE}
##these functions return named lists, one element for each provided EFO term
ckdSyns = getSynonyms("EFO:0003086")[[1]]
nsyn = length(ckdSyns)
##if odd add a blank
if( nsyn %% 2 ) { 
  ckdSyns = c(ckdSyns, "")
  half = (nsyn+1)/2
} else half= nsyn/2
outmat = data.frame(cbind(ckdSyns[1:half], ckdSyns[(half+1):(2*half)]))
colnames(outmat) = NULL
#ckdSyns
kable_styling(kbl(outmat, caption = "Synonyms for EFO:0003086"), bootstrap_options = "striped", full_width = FALSE)
```

```{r getstudyAcc, echo=FALSE}
ckdMatches = getMatchedTraits("EFO:0003086")
#length(ckdMatches[[1]])
ckdStudies = gwc_df[gwc_df$MAPPED_TRAIT_CURIE=="EFO:0003086",]
#ckdStudies$STUDY.ACCESSION

```

For clarity, there is one *document* associated with each EFO term. The term itself we call the *object*, which is augmented by a set of zero or more *synonyms* provided by EFO and then additionally
augmented by a set of zero or more *matches* which are GWAS study traits that have been mapped directly to the EFO term.

Searching for relevant GWAS studies using this corpus one first obtains all EFO terms where the tokenized search string matches at least one of the *object*, *synonyms* or *matches* for that term.  The desired GWAS studies are those that have either matched directly to that EFO term, or are inherited by that term. 

Suppose our search string had been *chronic kidney disease*, then we would have found 
the term EFO:0003086, which has as its text description *chronic kidney disease*. 
There were `r length(ckdMatches[[1]])` GWAS catalog studies that were mapped **directly** to this EFO term and `r efo_df["EFO:0003086", "Inherited"]` terms that were inherited, i.e. mapped to more specific EFO terms than EFO:0003086, which could be subtypes.

##FIXME: Vince here is a place where a picture of the EFO graph could be useful

## A search corpus based on GWAS Traits.

A dataframe, `gwc_df`, containing all GWAS studies is provided with the `gwasCatSearch` package.
For each study (e.g. GCST90083943), we report STUDY.ACCESSION: a unique GWAS Catalog identifier; DISEASE.TRAIT: the name the original researchers gave to the trait: MAPPED_TRAIT: which is the trait that the GWAS Catalog mapped to; MAPPED_TRAIT_URI: the URI for the mapped trait; and MAPPED_TRAIT_CURIE: the CURIE for the mapped trait. We note that for some studies there are multiple
rows, each one representing the mapping of DISEASE.TRAIT to unique EFO terms. In Table \@ref(tab:GWCexp1) we show the metadata associated with the study GCST90083943.

```{r GWCexp1, echo=FALSE}
knitr::kable(t(gwc_df[which(gwc_df$STUDY.ACCESSION=="GCST90083943"),]), align='lc', caption="Study GCST90083943", col.names="") |> kable_styling()

```

To create this searchable corpus we made documents, one for each GWAS study, by assembling the GWAS trait (DISEASE.TRAIT), the EFO mapped trait (Trait) and all EFO provided synonyms (Synonyms). In this case the ontology is being used to provide a standardized trait name and its synonyms. The relationships within the ontology are not used. The corpus can be searched and studies that match the search term retrieved. 

```{r GWCexp2, echo=FALSE}
##find all the studies that map to leukemia
htl = search_features(gwc_tc, query="leuk*")
htlf = addField2Hits(htl, gwc_tc)
tt = table(htlf$hits$field)
knitr::kable(tt, caption="Which part of the GWAS *document* matched the search term leuk*", col.names = c("Term", "Count"))  |> kable_styling()
```

As shown in Table \@ref(tab:GWCexp2) it is also possible to identify which of the three parts of the *document* described above have matches.
There were `r length(unique(htlf$hits$doc_id))` unique documents that
were identified. We can see that there were `r tt["DISEASE.TRAIT"]`
cases where a match was to the DISEASE.TRAIT, `r tt["Synonyms"]` matches
to synonyms provided by the ontologies, and `r tt["Trait"]` matches to
the text description used by the ontology for the trait. 

## Searching For Diseases

In this section we describe three different approaches that can be taken to
find GWAS studies that are associated with a particular term.  We will look for GWAS
studies associated with *infectious disease*.  One reason for this choice is that it is quite a challenging problem if one attempts it without using an ontology.  The main reason for that is that there are many infectious diseases and most of them do not have either *infectious* or *disease* in their name.  So without an ontology to guide us, we would need to have some dictionary of infectious diseases to start with. [FIXME: should we try to find one?]

### Strategy One: Use the EFO corpus


```{r tryChildren, echo=FALSE}
#first find the EFO term for infectious diseases
gg = grep("^infectious disease$", efo_df$Object, ignore.case=TRUE )
##efo_df[gg,]

##now look at all the more specific terms
z2 = EFOdescendants("EFO:0005741")
##length(z2)

efo_id = efo_df[z2,]
##d
## how many studies map directly to one of these terms
knitr::kable(t(table(efo_id$Direct)), caption="Number of studies mapping to descendants of EFO:0005741") |> kable_styling()
## one has 249 - and it is COVID, which makes some sense
# efo_id[which.max(efo_id$Direct),]

##drop those with 0 mapping to them
efo_ids = efo_id[efo_id$Direct>0,]

##below shows how to find the GWAS studies that map
dirMap = resources_annotated_with_term(efo_ids$Subject, include_subclasses=FALSE)
#dim(dirMap)
```

Here we describe how to use these tools start with a general type of disease, say infectious
disease, and then try to find all GWAS studies that report on that class of 
diseases. First we find the node/term in the EFO ontology that
corresponds to infectious disease. That turns out to be EFO:0005741. All infectious diseases should be
annotated in the EFO as descendants of that EFO term. You can use the `EFOdescendants` function to find all the descendants (there are `r length(z2)`). There were a total of `r efo_df[gg, "Direct"] + efo_df[gg, "Inherited"]` GWAS studies annotated at EFO:0005741 or one of its descendants. We find
that `r efo_df[gg,"Direct"]` were directly mapped to EFO:0005741 while `r efo_df[gg,"Inherited"]` where inherited. The term in Table \@ref(tab:tryChildren) with over 200 studies is `r efo_id[which.max(efo_id$Direct),"Object"]`

 FIXME: need some way to look at the text descriptions and see if they are sensible
 
### Strategy Two: Search based on the GWAS catalog corpus

For this search we start fairly broadly and search the `gwc_tc` for the term *infectious*. 
Recall that this will search the text descriptions of the GWAS trait as well as the trait and synonyms that correspond to the EFO term(s) this study was directly mapped to.

```{r IDgwctc}
qID = "infectious"
htsID = search_features(gwc_tc, query = qID)
htsID = addField2Hits(htsID, gwc_tc )
hitsIDasDT = hits2DTGWC(htsID, gwc_df, gwc_tc)
```

We can see where the matches occur. And see that the bulk have come from the EFO synonyms.

```{r checkIDmatches}
idTT = table(hitsIDasDT$field)
idTT
```


Now we can ask if this second approach found any GWAS studies that the first one did not.
And further look at those, to see if any are likely to be about infectious diseases. We
will also check to see if the first method found studies that the second approach did not.

```{r compID}
vv = setdiff(hitsIDasDT$Study, dirMap$STUDY.ACCESSION )
v2 = setdiff(dirMap$STUDY.ACCESSION, hitsIDasDT$Study)

DT::datatable(hitsIDasDT[match(vv, hitsIDasDT$Study),])
```

So there are `r length(vv)` studies that were found by the second method, that were not found by the first.  And `r length(v2)` studies found by the first, but not the second.  Which is probably what we hoped for. Many infectious diseases do not have the the words *infectious disease* as part of their name.

### Strategy three: Use the EFO ontology structure

Above we first started with the *infectious disease* in the EFO ontology and then found
all GWAS studies that mapped directly to either that node, or one of its descendants. We contrasted that with the strategy of searching the corpus based on the GWAS catalog studies for the terms infectious and disease.  A third approach is to search the EFO corpus for *infectious disease*
which we do in the code below

```{r otherterms}

qIDefo = "Infectious AND disease"
htsID2 = search_features(efo_tc, query = qIDefo)
hitsID2asDT2 = hits2DT(htsID2, efo_df, efo_tc)
##DT::datatable(hitsasDT2[1:100,], escape=FALSE, rownames=FALSE)

```

Now we can compare the EFO categories and see if we found anything new and also examine
the GWAS studies. A quick visual inspection of the EFO categories found by this method
and not the first, suggests that the first approach did seem to get everything useful.

```{r method3comp}
efoIDS3 = row.names(hitsID2asDT2)
setd3 = setdiff(efoIDS3, row.names(efo_id))
efo_df[setd3,"Object"]

```

We end this section by creating a Manhattan plot based on the studies that were identified using the first strategy. 

```{r getresources}

IDassoc <- granges_from_study(dirMap$STUDY.ACCESSION)
IDassoc2 = IDassoc[IDassoc$P.VALUE<1e-8,]

NROW(IDassoc2)
##how many are reported more than once...
sum(duplicated(IDassoc2$SNP_ID_CURRENT))
##and produce a plot
```

```{r IDmanh, plot=TRUE, fig.align='center', fig.width=7, fig.cap="Manhattan Plot for Infectious Disease", warning=FALSE, fig.height=5, fig.path="figure/"}
pp = variants_from_study(dirMap$STUDY.ACCESSION)
pp = pp[IDassoc$P.VALUE<1e-8,]

#pp = ggmanh::manhattan_data_preprocess(IDassoc2, pval.colname="P.VALUE")
simple_ggmanh(pp, y.label="-log10 p",
       label.colname = "MAPPED_TRAIT", pval.colname="P-VALUE")

##ggmanh::manhattan_plot(pp, plot.title="Infectious Diseases", y.label="-log10 p")
```

## Search for GWAS studies by Disease Location

In recent versions of the EFO the curators have begun to include disease location information.
This information is encoded using the UBERON ontology. To use this functionality you would first identify the UBERON CURIE that represents the location of interest.  In the code below we look for diseases associated with the lung.  FIXME: Vince - any sort of EFO graph that we could use to represent the disease categories associated with lung?

In the code below we identify the appropriate CURIE and then find all EFO terms that have been annotated with that term. FIXME - can there be sub-terms? and do we need some tools to roll up?
We next add a column to the metadata for the `efo_tc` corpus. This then allows us to subset the corpus and then we can carry out analyses based on the subset. Details on these capabilities are given in the @atteveldtCorpustoolsManagingQuerying2023 package vignette.

```{r uberon, warning=FALSE}
lungU = "UBERON:0002048"
lungEFO = grepl(lungU, efo_df$DiseaseLocation)


efo_tc$meta$Lung = lungEFO

efo_lung = efo_df[lungEFO,]
##FIXME - here it would be good to get some form of graph completion of the EFO labels

```

In the code below we subset first to the set of terms that have an association with the lung and then to the set of tokens that have more than 5 occurrances. And finally we create a word cloud based on the frequency of the tokens that remain.  These are based on occurance in the EFO corpus.

```{r subsetLung, warning=FALSE}
efo_tcLung = subset(efo_tc, subset_meta = Lung == TRUE)
efo_tcLung2 = subset(efo_tcLung, subset=freq_filter(token, min=5))
dtmLung = get_dtm(efo_tcLung, "feature")
csums = apply(dtmLung, 2, sum, na.rm=T)
sort(csums/sum(csums),dec=T)[1:10]
dtm_wordcloud(dtmLung)
```



## Exploring the GWAS metadata

Once we have a set of GWAS studies that we are interested in we can access the metadata
about which variants are implicated, where they are in the genome, the p-value and effect size to name a few.  This data is quite large and we use a SQLite database as the main tool for storing and accessing the data. The data are accessed via either `granges_from_study` if you want a `granges` object that can be integrated with a variety of Bioconductor @ tools for analyzing and plotting genomic data, or `variants_from_study` which returns a dataframe with 21 columns providing a set of metadata.

```{r from_studyfuncs}

tab = granges_from_study(acc="GCST007087")
tab3 = variants_from_study(acc="GCST007087")
names(tab3)
##do we care that these two are not the same dimenions?  

```

Then one can also look at whether the same genes are implicated across
studies. Note that there are #r NROW(assoc3)# associations that are
plotted. The interested reaader can explore a number of different
aspects of these mappings.

```{r lookatHits, eval=FALSE}
sort(table(assoc3$MAPPED_GENE),dec=TRUE)[1:10]
tt = sort(table(assoc3$MAPPED_TRAIT_CURIE), dec=TRUE)
tt[1:10]
```

Then look at some of the traits that show up often. The top two hits are
for lymphocyte counts. Out of the top 10, only one seems to actually be
an infectious disease. The others are autoimmune, COPD, and measurements
of lymphocytes.

```{r maptotraits, eval=FALSE}
efo_df[names(tt)[1],]
efo_df[names(tt)[2],]

```



## Challenges

While the GWAS catalog has provided mappings to EFO there remain some substantial challenges
for those attempting to use it to find and organize their searches.  These include vague EFO assignments where hundreds of phenotypes have been mapped to an EFO term, papers that give rise to thousands of GWAS studies, inaccuracies and inconsistencies in the EFO, and a lack of clarity on the explicit rationale for mapping a trait to an EFO term.  We give examples of each of these.

### Vague Study Definitions

Sometimes the EFO terms used are very generic and hence don't provide much resolution of phenotypes.
By that we mean that they have too many studies that map to them and hence are unlikely to be informative about the sort of specific questions we might want to address. These are
relatively easy to find. Using a cutoff of 500 GWAS studies we obtain the set of EFO terms that have more than 500 GWAS studies that have been directly mapped to them. 


```{r studies2EFO, echo=FALSE}

manystudies = efo_df[efo_df$Direct>500,]
knitr::kable(manystudies[,c(1,2,5,6)], row.names=FALSE, caption="List of EFO terms with more than 500 Studies", label="tab:xx", full_width=FALSE)

```

One can then get a sense of the actual traits that were mapped to these by using one querying the GWAS catalog dataframe. We will look at EFO:0004725, which is *metabolite measurement*. We identify
all GWAS studies that were directly mapped to EFO:0004725 and then examine the GWAS trait strings.
In are we see that this EFO term is really a catch-all for a wide variety of chemical measurements and hence unlikely to be helpful providing insight into the commonality of terms that were mapped directly to it. 

##FIXME: Ask Rafael about other ideas here - maybe there is some sort of diversity index that could be computed per EFO term that says something about how related the terms mapped there are

```{r vagueTraits, echo=FALSE}
metabTraits = gwc_df[gwc_df$MAPPED_TRAIT_CURIE=="EFO:0004725", "DISEASE.TRAIT"]
zz = table(metabTraits)
knitr::kable(sort(zz,dec=T)[1:10], row.names=FALSE, caption="GWAS disease traits mapping to EFO:0004725")
```

### Mapping not so great

In this example we are examine kidney function traits, and in particular we
will search for *chronic AND kidney AND disease* in the EFO-based corpus. This string should find EFO terms that have matchs to all three words that are close to each other.

```{r gwasCatPubMed, echo=FALSE}
q1 = "chronic AND kidney AND disease"
hts = search_features(efo_tc, query = q1)
hts = addField2Hits(hts, efo_tc )
hitsasDT = hits2DT(hts, efo_df, efo_tc)
```

```{r DT, fig.cap="CKD EFO Hits", echo=FALSE}
knitr::kable(hitsasDT[1:10,], escape=FALSE, row.names=FALSE, caption="CKD EFO Hits (first 10)")
##DT::datatable(hitsasDT[1:100,], escape=FALSE, rownames=FALSE, caption="CKD EFO Hits (first 100)")
```

Surprisingly we find that there were `r nrow(hitsasDT)`
different EFO terms that have a hit. This seems very large to
understand what is happening we start by looking 
code chunk we explore some of the issues. If we consider the EFO node for
*kidney disease* we anticipate that all different forms of kidney
disease will be descendants of that node. And we expect that all `r nrow(hitsasDT)`
hits will map to some one of those nodes.

```{r ckdEFO, echo=FALSE}
kdEFO = grep("^kidney disease$", efo_df$Object, ignore.case = TRUE)
efo_df[kdEFO,]
ackd = EFOdescendants("EFO:0003086")
```

There are only `r length(ackd)` EFO terms that are more specific than the term *kidney disease* and the total number of GWAS studies that mapped directly to EFO:0003086 or were inherited we only get `r efo_df["EFO:0003086","Direct"] + efo_df["EFO:0003086","Inherited"]` GWAS studies.  This suggests that there are a very large number of studies where the GWAS trait contains the phrase *chronic kidney disease* that were not mapped to any EFO kidney disease. 


```{r msmt, echo=FALSE}
msmtdesc = EFOdescendants("EFO:0001444")
dtandmsmt = intersect(msmtdesc, row.names(hitsasDT))
length(dtandmsmt)

```
A little more exploring suggests that many of the terms are actually mapped to the descendants of the *measurement*, EFO:0001444, node. Specifically we find that there are `r length(dtandmsmt)` EFO terms that were identified by our search which align to measurements, a number far larger than those that align to disease.  Which makes searching for terms related to any specific disease somewhat more challenging, as the organizational principles for this part of the EFO are not suitable for finding related diseases and it appears the analyst is left with string matching as the only viable option.

We constructed our documents by combining the name of the EFO trait (Object), the EFO synonyms (Synonyms) and all GWAS traits that were directly mapped to that EFO trait (Matches) into one *document*.  We can query the corpus to determine which of these three parts any match was made to.
the matches to *chronic kidney disease* are occurring. For example, for `r hitsasDT$EFO[1]`
the string *chronic kidney disease* was matched only to other GWAS
catalog traits (since only matches were found).

```{r why7K, echo=FALSE}
knitr::kable(table(hitsasDT$field))
```

This means that the bulk of the EFO terms are showing up due to finding
the text string in the name of some other GWAS catalog trait and 
that trait was not mapped to any EFO label related to kidney disease. 


We next query the GWAS catalog traits.
The dataframe `gwc_df` contains the names of all traits in the `DISEASE.TRAIT` column.

```{r gwcmetadtable, echo=FALSE}

gg = grep("chronic kidney disease", gwc_df$DISEASE.TRAIT, ignore.case=T)


gl= grepl("chronic kidney disease", gwc_df$DISEASE.TRAIT, ignore.case=T)
gwsub= gwc_df[gl,]
r1 = grep("measurement",gwsub$MAPPED_TRAIT,  ignore.case = TRUE)
#length(r1)

##gwc_df[gg[1:10], "DISEASE.TRAIT"]
ckdPMIDs = GWCATID2PMID(unique(gwc_df$STUDY.ACCESSION[gg]))
pmtab = sort(table(ckdPMIDs), dec=TRUE)
#pmtab[1:10]
```

And now we see that there are in fact `r length(unique(gwc_df[gg, "STUDY.ACCESSION"]))` terms that have been labeled this way. Further, if we look at the PUBMEDIDs associated
with these traits we see that there are `r sum(pmtab>1000)` papers that
report more than 1000 traits. Given the PubMed IDs it is easy to find
the papers. The paper associated with PMID 35870639 which reports more than 
6,700 studies is titled: Identification of 969 protein quantitative trait loci in an African American population with kidney disease attributed to hypertension @surapaneniIdentification969Protein2022.


We end this example by creating a Manhattan plot based on the studies that contain the string *chronic kidney disease*.

```{r ckdmanh, plot=TRUE, fig.align='center', fig.width=7, fig.cap="Manhattan Plot for CKD", warning=FALSE, fig.height=5, fig.path="figure/"}
vars = variants_from_study(gwc_df$STUDY.ACCESSION[gg])
##drop those with a p-value > 1e-8,
vars = vars[vars$`P-VALUE` < 1e-8, ]
simple_ggmanh(vars, y.label="-log10 p", plot.title = "Chronic Kidney Disease",
       label.colname = "MAPPED_TRAIT", pval.colname="P-VALUE")

```

### Bad synonyms

Another potential issue is that the EFO provided synonyms might be inaccurate. Further investigation of those issues is also recommended. FIXME: maybe Rafael has some ideas.


#### Other Hierarchies

While much of the EFO of interest is organized around *diseases* (EFO:0000408) there is also
a *measurement* (EFO:0001444) node, with about 8,700 terms nested under it. These tend to be measurements of some of the very same diseases that are arranged under the *disease*  node. Unfortunately, there is no easy way to go between the two. They are not related in the EFO hierarchy and we have been unable to find an organizing principle.  Users likely will want to search under both of these high level nodes for matches to any disease, gene or phenotype of interest.   For example, in our infectious disease example we see that there are nodes 
*susceptibility to infectious disease measurement* (EFO:0008422) and
*infectious disease biomarker* (EFO:0006843) which may also be relevant.
FIXME: report the number of studies that. align under the measurement node


 Messing around with some NLP stuff - not sure I can see how to use it though
 need some way of mapping to a disease dictionary....
 
```{r spacyr, eval=FALSE }
##this is our dictionary - 
InfDisease = scan("InfDisease.txt", what="", sep="\n")
z2 = EFOdescendants("EFO:0005741")
InfDfromEFO = efo_df[z2,"Object"]
names(InfDfromEFO) = z2
library("spacyr")
spacy_initialize()

nn = paste0("d", 1:length(InfDisease))
names(InfDisease) = nn

nn = paste0("e", 1:length(InfDfromEFO))
names(InfDfromEFO) = nn
combdata = c(InfDfromEFO, InfDisease)


parsedtxt <- spacy_parse(combdata)

s1 = grepl("e", parsedtxt$doc_id)

sub1 = parsedtxt[s1, ]
sub2 = parsedtxt[!s1,]

nn = paste0("d", 1:length(InfDfromEFO))
names(InfDfromEFO) = nn
ptxtEFO = spacy_parse(InfDfromEFO)

vv = tolower(ptxtEFO$lemma)
v1 = tolower(parsedtxt$lemma)

d1 = setdiff(vv, v1)




```

Next we have tried using the scispacy tools.
Rafael provided these files:
The resulting table for each file is structured as follows:
- ID: an autogenerated UUID is attached to each input line in the given file
- InputText: the input line of text from the file
- Entity: the Entity that spacy identified in the InputText
- CUI: the UMLS unique concept identifier mapped to the Entity
- Label: the UMLS (preferred) label for the concept
- Definition: the definition for the concept in UMLS
- Synonyms: the synonyms of the concept in UMLS
- Score: the score/confidence of the mapping between entity <-> UMLS concept

There are sometimes multiple output lines for each input. That means that scispacy identified
a set of concepts in the input.  Some are probably of the form (from the repeated entities)
infectious disease; human; primary; commensal; infection; opportunistic; bacterial; disease;
fungal; iatrogenic; bacterimia,
which are somewhat generic

A second kind of duplication is where the same entity appears multiple times. The most common
one is infectious disease.



Things that did not map from EFO to UMLS - some fixes suggested below
setdiff output...for now:
> setdiff(efoIp, efoInfDss$InputText)
 [1] "typhus"                          "bacteriemia"                    
 [3] "kuru"                            "dysentery"                      
 [5] "scrapie"                         "perichondritis of auricle"      
 [7] "Kluver-Bucy syndrome"            "chickenpox"                     
 [9] "Chikungunya encephalitis"        "spotted fever"                  
[11] "influenza"                       "BK-virus nephropathy"           
[13] "molluscum contagiosum"           "common wart"                    
[15] "whooping cough"                  "onchocerciasis"                 
[17] "Acanthamoeba infectious disease" "rhinosporidiosis"               
[19] "screw worm infectious disease"   "Klebsiella infectious disease"  
[21] "primary systemic mycosis"        "Rhabditida infectious disease"  

Fixups:  typhus is a synonym for Typhoid fever (in the UMLS mapping - so perhaps we
should use those), then kuru is missed but Kuru is found (so capitalization issues)
Chikungunya is there, but not encephalitis, which is just a modifier,
chickenpox - Chickenpox
Acanthamoeba Infection is there - that seems like a horrible miss
Acanthamoeba keratitis...mostly Acanthamoeba seems like the thing to match on...
then 
EFO:  digestive system infectious disorder which is MONDO:0043424 - and seems important
in the EFO world as an organizer, but probably not something to match on.

"Klebsiella infectious disease" in the other list is: Klebsiella Infection. So something that
the NER stuff should have handled, I would have thought

molluscum contagiosum - is common more than 200K cases per year, so it should be there somewhere


Do we have a leaves function - maybe we want to try and map the leaves (and worry less about
non-leaves)

We have CUI labels for lots of the EFO terms, but these same labels are not found in the other list - that might not be a bad thing.


```{r scispacyoutput, eval=FALSE}


efoInfDss = read.delim("data/InfDfromEFO_scispacy_entities.tsv", sep="\t", head=T)
IDss = read.delim("data/InfDisease_scispacy_entities.tsv", sep="\t", head=T)
un1 = unique(efoInfDss$CUI)

mm2 = match(efoInfDss2$CUI, IDss$CUI)
sum(is.na(mm2))


##look at matching between our two lists - which did not seem to work that well,
##but we did match to the CUIs for almost all of the elements

tsub = efoInfDss2[!is.na(efoInfDss2$EFOID),]
mmx = match(tsub$CUI, IDss$CUI)
matchedEFO = ifelse(is.na(mmx), FALSE, TRUE)
IDleafs = isLeaf(tsub$EFOID)
table(matchedEFO, IDleafs)
chisq.test(matchedEFO, IDleafs)

efoInfDss$Entity[duplicated(efoInfDss$Entity)]

## look at some terms that map to CUIs but are not found in the InfDss list

length(mm2)
ts2 = efoInfDss2[is.na(mm2),]
## if we look at the duplicated terms
sort(table(efoInfDss$Entity[duplicated(efoInfDss$Entity)]), dec=TRUE)[1:20]

dupID = efoInfDss$ID[duplicated(efoInfDss$ID)]
efoInfDss[efoInfDss$ID %in% dupID[1], 1:4]
efoInfDss[efoInfDss$ID %in% dupID[2], 1:4]
efoInfDss[efoInfDss$ID %in% dupID[2], 1:5]

##put in the EFO terms
mmEFO = match(InfDfromEFO, efoInfDss$InputText)
efoInfDss2 = efoInfDss
Ind2 = InfDfromEFO[!is.na(mmEFO)]
mmEFO = mmEFO[!is.na(mmEFO)]
efoInfDss2[mmEFO, "EFOID"] = names(Ind2)

##now we can try to figure out if the Leafs match more often than non-leaf nodes

##now ask how many of the UMLS terms attached to EFO are attached to the ID list

efoU = unique(efoInfDss$CUI)
idU = unique(IDss$CUI)

ss = setdiff(efoU, idU)
ss2 = setdiff(idU, efoU)
length(ss)

length(ss2)


##check to see what is different in the EFO labels I sent vs got back
efoIp = scan("../InfDfromEFO.txt", what="", sep="\n")
##see what is in the EFO that didn't map
##scrapie is non-human
setdiff(efoIp, efoInfDss$InputText)
grep("Typh", IDss$InputText, ignore.case=T)
IDss[741,1:6]
names(IDss)
IDss[741,7:8]

##the mapping scores seem very high - so they can be used
hist(efoInfDss$Score)
hist(IDss$Score)
grep("uru", IDss$InputText, ignore.case=T)
IDss[89,1:5]
IDss[416,1:5]
grep("riemia", IDss$InputText, ignore.case=T)
grep("crapie", IDss$InputText, ignore.case=T)
grep("crapie", IDss$Synonyms, ignore.case=T)
grep("Jakob", IDss$InputText, ignore.case=T)
IDss[762, 7:8]
IDss[762, 1:5]
IDss[763, 1:5]
IDss[763, 7:8]
grep("gunya", IDss$InputText, ignore.case=T)
IDss[144,1:6]


```

## Citations



