---
title: "gwasCatSearch Quick Start and More"
author: "Robert Gentleman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gwasCatSearch Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
format: html
editor: visual
---

## Introduction

```{r echo=FALSE, warning=FALSE, message=FALSE}
library("gwasCatSearch")
```

The gwasCatSearch package provides functionality to search for different phenotypes within the GWAS catalog (<https://gwas.mrcieu.ac.uk>) database. We used the mappings provided by the GWAS catalog for phenotypes to the EFO ontology (<https://www.ebi.ac.uk/efo/>). We then created a SQLite database that contains (among other things) a table that has the linearized form of the ontology. Annotations can be mapped directly to a term, or they can be inferred from any child term. Then given any EFO ontology label users can query the database to determine which, if any, GWAS catalog phenotypes have been mapped to that EFO term.

While helpful, we then have the challenge of how to help a researcher find an EFO label that corresponds to a topic they are interested in. To help solve that challenge we took all the EFO term labels and created a searchable corpus using the corpustools R package (<https://cran.r-project.org/web/packages/corpustools/index.html>) so that users can start with ordinary words and then find EFO terms that might be of interest. The output of this search can be accessed programmatically or we provide a conversion to a datatable that can be viewed, sorted and searched.

## Searching for EFO terms with specific words

In the first example below we search for any *document*, which in our case is just the text label for an EFO ontology term that contains a match to either *granulo\** or *rheum\**. We named these two arguments **Granuloma** and **Rheumatic**. The `summary` function tells how many terms matched and in how many *documents*. And then we created a table using `hitsasDT` and then rendered that using `datatable`. The output is searchable. Under the column labeled **Granuloma** are the actual words that were matched to *granulo\** and under the column labled **Rheumatic** are the words that matched to *rheum\**.

```{r}
hits <- search_features(efo_tc, query = c("Granuloma# granulo*", "Rheumatic# rheum*"))
summary(hits)
hitsasDT <- hits2DT(hits, efo_df, efo_tc)
DT::datatable(hitsasDT, escape = FALSE, rownames = FALSE)
```

In the next code chunk we search for any EFO labels that match to both *granulo\** and *rheum\**. There are none.

```{r}
ht2 <- search_features(efo_tc, query = "granulo* AND rheum*")
summary(ht2)
```

## Searching for matches from GWAS Catalog hits.

Here we describe how to search for GWAS catalog hits that match a certain EFO identifier. For our example we use the EFO term [EFO:0007987](#0) which is labeled as *granulocyte count*. To do this we call the `resources_annotated_with_term` function. We provide the term identifier and then either ask only for the resources annotated specifically with the provided term (`ematch`) or for all those that either match directly, or that match to a child of the term provided (`allmatch`). There are 4 that match directly and another 106 that match for some children. We use a `datatable` so the reader can explore the first 20.

```{r}
ematch <- resources_annotated_with_term("EFO:0007987", include_subclasses = FALSE)
dim(ematch)
allmatch <- resources_annotated_with_term("EFO:0007987", include_subclasses = TRUE)
dim(allmatch)
DT::datatable(head(allmatch, n = 20))
```

## Description of the Database

At some point these details will go into their own document with some additional information on the database schema. But for now we describe how we created the mappings here. This is definitely just a first pass at the approach and since we make our mappings public, interested readers could help by pointing out particularly poor or inaccurate mappings so we can improve our approach.

We mapped OpenGWAS phenotypes to EFO in the following way. The inputs to `text2term` were a table containing the OpenGWAS metadata from 2022-01-25 and the EFO ontology v3.43.0. We configured `text2term` to include only mappings with a score above our minimum threshold (`min_score=0.6`, in a $$0,1$$ scale where 1=exact syntactic match), and to compute only the highest scored mapping for each trait in the metadata (`max_mappings=1`). We use the [TFIDF](https://en.wikipedia.org/wiki/Tf%E2%80%93idf) mapper provided by `text2term` (`mapper=Mapper.TFIDF`), which computes TFIDF-based vector representations of traits and then uses cosine distance to determine how close each trait is to each ontology term (by considering ontology term labels and synonyms encoded in EFO). Finally we exclude terms that have been marked as deprecated (`excl_deprecated=True`) such that we only map to terms that are current and expected to be in EFO's future releases.

EFO contains terms and relationships between terms that exist in external ontologies such as MONDO, ChEBI, etc. Since our goal is to map phenotypes to appropriate terms in ontologies, if they exist, we further configured text2term to only map to terms from ontologies that describe phenotypes: EFO itself, the Monarch Disease Ontology (MONDO), the Human Phenotype Ontology (HPO), and the Orphanet Rare Disease Ontology (ORDO). To do this, we use the parameter `base_iris` in text2term which limits search to terms in the specified namespace(s), which we have set as follows: '<http://www.ebi.ac.uk/efo/>', '<http://purl.obolibrary.org/obo/MONDO>', '<http://purl.obolibrary.org/obo/HP>' and

$$ORDO$$(<http://www.orpha.net/ORDO>).

min_score=0.6,\
\# minimum acceptable mapping score\
mapper=Mapper.TFIDF,\
\# use the (default) TF-IDF-based mapper to compare strings\
excl_deprecated=True,\
\# exclude deprecated ontology terms max_mappings=1,\
\# maximum number of mappings per input phenotype
