---
title: "Working with Ontologies"
author: "Robert"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gwasCatSearch Working with Ontologies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gwasCatSearch)
```

## Using the ontologies
The `efo_oi` data set provides access to information about the EFO ontology that has been organized along the lines of the ontologyIndex package from CRAN.

```{r }
efo_oi$name["EFO:0000095"]
efo_oi$children["EFO:0000095"]
efo_oi$parents["EFO:0000095"]
```
One of the parents of `EFO:0000095` is `MONDO:0004095` and we can see how many children it has
by using the `children` table in the ontology index.
If we use the `[[` accessor then we just get the vector of children. If instead we use the `[` accessor then we would get a list that is named by the input vector of terms where each element is the vector of children for the corresponding EFO term.
```{r}
x1 = efo_oi$children[["MONDO:0004095"]]

x1
```

Now we will interact with the DB to get at the mappings to different terms.

```{r dbmappings}

 query = paste0("SELECT * FROM efo_labels WHERE Subject IN ('",
                paste(x1, collapse = "','"),
                "')")
 mappedans = DBI::dbGetQuery(gwasCatSearch_dbconn(), query)
 mappedans$Direct
 mappedans$Inherited

```

Now we can look at some plotting for the graph and manipulations of the EFO tables
```{r graphPlot}
tags = c("EFO:0003884", "EFO:0004531", "EFO:0005208", "EFO:0009909", "EFO:0021409", "EFO:0021433")
##assign the next one to get back the graph structure that is plotted
xt = plotTags(tags)
nn = graph::nodes(xt)
##nodes have long names - suitable for printing we just want the CURIEs
v1 = strsplit(nn, "\n")
nodeEFO = sapply(v1, function(x) x[length(x)])
subs=efo_df[nodeEFO,c("Direct","Inherited")]
subs
```